---
title: "Exercise 3.2: Displaying all data"
output: html_document
---

Show the dots behind the data. 

The quick way. We can use packages that do this for you.

Combining different elements on the same plot can be harder  

# solving the overplotting problem

# Overplotting problem (position layer)
# from RFDS examples
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")

ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
You can jitter the data.

```{r}
# Increase the 
ggplot(finch_data, 
  aes(x = species, y = Beak_length.mm)) + 
  geom_point() 
```

### Subtitle

Description

```{r}

```

Reflection Questions 

Key Takeaways

Additional Resources
