---
title: "Exercise 3.3: Publication quality plots"
output: html_document
---

Here are some considerations to produce publication-ready plots: 
- Submission standards, such as file types and dots per inch.
- Colour blindness
- Font size readable on a two column vs one column plot in a paper

We'll continue with the finch data. Run this again or skip if you follow from exercise 2.

```{r}
# The working directory needs setting to day1/03_session 

library(tidyverse) 

# let's load the Finch data, it is actually four datasets
beak_t1<- read.csv("./data/finch_beaks_1975.csv")     # peak measurements in1 1975 for scandens / fortis
beak_t2<- read.csv("./data/finch_beaks_2012.csv")     # peak measurements in1 2012 for scandens / fortis

beak_t1 <- beak_t1 %>% mutate(year = 1975)
beak_t2 <- beak_t2 %>% mutate(year = 2012)

# Combine the datasets using bind_rows (dplyr::bind_rows())
finch_data <- bind_rows(beak_t1, beak_t2)

```

# Colour blindness

We tend to forget that colour deficiencies (colour blindness of some type) are extremely prevalent. People can then have difficulties in telling the difference between conditions.

```{r}
# We need the package ggthemes exptending ggplot capabilities
install.packages("ggthemes")
library(ggthemes)

ggplot(finch_data, aes(x = Beak_length.mm, y = Beak_depth.mm, colour = species)) + 
  geom_point()+
  scale_colour_colorblind()  # Apply colorblind-friendly palette

# We now have a plot where differences are defined by brightness mostly, and color
# However, We have the problem that the contrast of the orange dots is not very good
# fix somewhat by adding the theme_classic() and increasing dot sizes

```

# Saving publication-ready plots

Use ggsave to increase dpi from base R level (72 dpi) and to defining figure dimensions. It is common to increase to 300-600 dpi for high-quality graphs. 

For instance, last time I checked the journal Nature will display 3.5 in (89 mm) of a single column figure and 7.2 in (183 mm) of a double column figure. PNG and PDF formats are very common requests for production.

For a publication-quality figure, a common practice is to use a font size of 10 to 12 points for axis labels and legends. This size ensures that the text is legible when printed or viewed on a screen. As a rule of thumb make the text within the figure as big as the surrounding text if not bigger.

Next we are saving a plot meant for a 3.5-inch width (single-column).

You are faced with the issue of specifying the aspect ratio in inches? 

You need first to show the plot in the clipboard. For that you would need to execute in the Console (e.g. copy and paste)

1. Go to the "Plot" tab in the bottom-right window. Go to "Export" > "Copy to clipboard"

2. Grab the bottom right corner of the image to change the aspect ratio. See the numbers on top (pixels) change. Make sure "Maintain aspect ratio" is deselected 

3. Find an aspect ratio (ratio of width to height) that works for you. You could insert into a word file and print to get a good idea of what it would look like in one column of a paper. Copy the pixel values for width and height. Those determine the aspect ratio.

The aspect ratio is width / height You know the figure should occupy 3.5 inches. You multiply by the aspect ratio to get the height in inches.

For example, if the width is 3.5 inches and the aspect ratio is 1.4, the height would be 3.5 / 1.4 = 2.5

4. Adjust font size family if too tiny

```{r eval=FALSE, include=FALSE}
# plot a production-ready figure (let's not be too finicky at this point)
ggplot(finch_data, aes(x = Beak_length.mm, y = Beak_depth.mm, colour = species)) + 
  geom_point(size=1) +  # Add points for each observation
  theme_classic(base_size = 10) +  # Use a classic theme for a clean look
  labs(
    x = "Beak Length (mm)",  # Label for x-axis
    y = "Beak Depth (mm)",    # Label for y-axis
    colour = "Species"        # Legend title
  ) +
  theme(
    # change the size of individual elements by uncommenting this if you want
    #axis.title.x = element_text(size = 12),  # X-axis title font size
    #axis.title.y = element_text(size = 12),  # Y-axis title font size
    #axis.text = element_text(size = 10),  # Axis text font size
    #legend.title = element_text(size = 12),  # Legend title font size
    #legend.text = element_text(size = 10)  # Legend text font size
  )

# Fill the blank

# Specify the size in pixels for the height and width (as determined above)

# save a png file
ggsave("figures/finch_data_hr.png",dpi=600, height=___, width=__, units="in")

# note you are not getting exactly what you are seeing on the preview, when it comes to dot size
# open and zoom on the figure with the Files tab
# see the difference with different values of dpi
# adjust the size of dots and base_size to optimal values

# save a pdf 
ggsave("figures/finch_data_hr.pdf", device="pdf",dpi=600, height=__, width, __)

```

# To cap the day: 

Make the fanciest possible plot of the scandens beak length by year, save it and upload it. Winner will be judged based on aesthetic quality and application of the principles we have covered. You can use AI (e.g. Copilot) to make as sophisticated a plot as you want and bring this bird data to life; you are not limited to what we have covered or the packages we have used.

Upload the file using this [link](https://forms.gle/NYreXR7rvZc2rJzH8). 

Name the file: (Your Name)_(Your Surname)_pretty_finch.png

# Key Takeaways

File Types: Use PNG or PDF for high-quality outputs.

Resolution (DPI): Increase DPI to 300–600 for publication.

Font Size: Use 10–12 points for axis labels and legends.

Aspect Ratio: Maintain a suitable width-to-height ratio (e.g., 3.5 inches wide for single-column figures).

Additional resources

Edward Tufte book: Tufte, E. R., & Graves-Morris, P. R. (1983). The visual display of quantitative information (Vol. 2, No. 9). Cheshire, CT: Graphics press.

You can see a summary [here](https://benjaminleroy.github.io/pages/blog/public/post/2018/05/16/review-of-tufte-s-the-visual-display-of-quantitative-information/)

Stephen Kosslyn's book: Kosslyn, S. M. (2006). Graph design for the eye and mind. OUP USA.

You can see a very short summary here [here](https://civilstat.com/2012/11/graph-design-for-the-eye-and-mind-stephen-kosslyn/)