---
title: "Exercise 3.2: Displaying All Data"
author: David Souto
output: html_document
---

## Show the dots behind the data

The quick way to visualize data effectively is to use packages that help manage overplotting. Combining different elements on the same plot can sometimes lead to clutter, but there are strategies to solve this problem.

Run the following chunk if needed to get started.

```{r}
# The working directory needs setting to day1/03_session 

# let's load the tidyverse, which holds ggplot
library(tidyverse) 

# let's load the Finch data, it is actually four datasets
beak_t1<- read.csv("./data/finch_beaks_1975.csv")     # peak measurements in1 1975 for scandens / fortis
beak_t2<- read.csv("./data/finch_beaks_2012.csv")     # peak measurements in1 2012 for scandens / fortis

beak_t1 <- beak_t1 %>% mutate(year = 1975)
beak_t2 <- beak_t2 %>% mutate(year = 2012)

# Combine the datasets using bind_rows (dplyr::bind_rows())
finch_data <- bind_rows(beak_t1, beak_t2)

```

### Solving the Overplotting Problem

To address overplotting, the issue we have with dots being on top of each other, we can use the `position` argument in `geom_point()` to jitter the points, making them more distinguishable. 

```{r}
# We can show a boxplot of beak lengths by species and see the problem straight away
ggplot(finch_data, aes(x=species, y=Beak_length.mm))+
  geom_point()
```
We could sort this problem by jittering the dots by some amount. It is not cheating if we jitter along the x-axis, as we are not altering beak length. 

```{r eval=FALSE, include=FALSE}
# Let's try jittering along the by using position 
ggplot(finch_data,aes(x = species, y = Beak_depth.mm)) + 
  geom_point(position = "jitter")

# Fill the blanks

# by default this is a lot of jitter, also we want an horizontal jitter, we can adjust that with postion_jitter()
# adjust the parameters to something that feels right
ggplot(finch_data,mapping = aes(x = species, y = Beak_depth.mm)) + 
  geom_point(position = position_jitter(width=__,height=__))

# We 
# Bar plot with transparency for clarity ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + geom_bar(alpha = 1/5, position = "identity")

```

# Combining elements

We are not restricted to just one type of geometry, we can combine them.

```{r}

# Add a geom_boxplot() to this graph (remember a + to indicate we are adding a layer)
ggplot(finch_data,aes(x = species, y = Beak_depth.mm)) + 
  geom_point(position = "jitter")+
  geom_boxplot()

# See what happens if we invert the order of the geom_'s

```
# Adding averages and error bars

With the stat_ layer we have the option of calculating several summary statistics within ggplot. That way we can display error bars on top of dots.

```{r}
ggplot(finch_data,aes(x = species, y = Beak_depth.mm)) + 
  geom_point(position = "jitter")+
  # we specify we want a standard deviation
  stat_summary(fun.data = "mean_sdl", 
               # specify we want an errorbar geometry
               geom = "errorbar", colour="red",size=1, width=0.2)+

# we have now created the ugliest error bar possible: make it pretty with the parameter size and colour in stat_summary; 
# add another summary corresponding to the mean on top, note a subtlety here, you specify fun="" and not fun.data="" because it is a single value, specify size and colour again

```

# Mixing datasets

What if we have done the work of summarizing the data already with say `dplyr` in some specific way we like? How can we mix the observation-level data with the averages?

```{r}
# Mixing datasets

# say we have a data summary for the finch dataset
# we know how to do this
beak_summary<- finch_data %>% 
  group_by(species) %>% 
  summarise(mean = mean(Beak_length.mm), # calculate mean
              sd = sd(Beak_length.mm))   # calculate standard deviation

print(beak_summary)  

# Now combine raw data with summarized data in a ggplot
ggplot(finch_data, aes(x = species, y = Beak_length.mm)) +
  # We start by adding raw data points with jitter
  geom_point(position = position_jitter(width = 0.2), 
             alpha = 0.5, size = 2, color = "blue")+
  # Add mean points for the summarized data
  # sublety here: we have to be more specific, omitting 'data', as we did for ggplot() won't work
  geom_point(data = beak_summary, 
             aes(x = species, y = mean), 
             size = 3, color = "red")
```
# Violin and dot plots

Rain and violin plots can be used to efficiently show dot distributions, including with large amounts of data. To do so with `ggplot` we can use `stat_halfeye`, `geom_violin`.

```{r}
ggplot(finch_data, aes(x = species, y = Beak_depth.mm)) +
  # Add a density plot
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.2,
    point_colour = NA
  ) +
  # Add boxplot
  geom_boxplot(
    width = 0.12,
  ) +
  # Add dot plot
  stat_dots(
    side = "left",
    justification = 1.1,
    binwidth = NA 
  )
```

Another version uses `geom_violin`, producing a proper violin plot, basically a mirrored density plot.

```{r}

# Fill the blanks
ggplot(finch_data, aes(x = species, y = Beak_depth.mm)) +
  # Add half-violin plot, 
  geom_violin()+
  geom_boxplot()

# This plot doesn't win any beauty contests; use 'width' to adjust the width of the boxplot
# change the smoothing bandwidth of the violin plot with 'bw', or 'adjust' (a factor of default value), try different values

```
# Level 2 (optional): Advanced Visualization Using Specialized Packages

You would have noticed that combinations of dot / violin plots are a bit fiddly to handle with base ggplot. Package `ggdist` facilitates some of this. 

The package has its own cheat-sheets and documentation](https://github.com/mjskay/ggdist/tree/master).
              
Key Takeaway: 

When combining multiple statistical elements, focus on clarity and avoid visual clutter. 

Use transparency and appropriate positioning to ensure all elements are visible and interpretable.