---
title: "Exercise 1.1: Workshop packages (20 mn)"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

We are going to look explore some of the characteristics of the packages
we installed, in particular those are are part of the *tidyverse*

If you didn't install this already:

-   Set up the working directory to the day-1 folder: click on the
    "Session" tab (on top) then "Set working Directory" then "Choose
    Directory" then "Open".

-   Run source("../../Pre_Package_Install.R") if you didn't install the packages
    yet.

```{r}

# we are telling R to source (run) the file that is two folders up with "../.."
source("../../Pre_Package_Install.R")

```

### Start by loading the tidyverse

The tidyverse is a collection of R packages designed for data science
that share a common philosophy and design. In this worksheet, we'll
explore: - What packages are included in the tidyverse - How to handle
package conflicts - Compare tidyverse with base R approaches

```{r}
# Use Ctrl+Enter (Cmd+Enter) to run line-by-line, rather than by chunk
library("tidyverse")

# find out how to cite in a publication (this format can be imported to Zotero for instance)
citation("tidyverse")

# Optional: Find out general information about the package with library(help="tidyverse"), find out what a meme is with library(help="meme")

```

Look at the message that appears when loading tidyverse. Which core
packages are loaded by default? Take note of conflicts. Those are
functions within one of the packages that have the same name as other
loaded packages (e.g. part of base R)

### Managing package conflicts

There are different ways to manage naming conflicts. Or you could be
explicit about those functions by using the "package_name"::"function
name" syntax. You could also install the 'conflicted' package, which
will throw an error when 'overloaded' functions are found.

```{r}
# Run this chunk and observe the output 
filter(mpg, cyl==4) # we are filtering the pre-loaded fuel economy data frame by number of cylinders

# Do the same using "filter" within the dplyr package by using the :: syntax, 

# Do the same using the "stats" package  

# Optional: install the conflicted package (syntax: install.packages(""), then load with library()) and observe the output when running the first line

# Optional: Can you guess what dplyr is about by browsing through package functions with the :: syntax? 
```

### Data filtering

Let's see how we filter data in the tydiverse by selecting cars with 6
cylinders:

```{r}
# Base R approach 
base_r_way <-mtcars[mtcars$cyl == 6, ]

# Tidyverse approach 
tidyverse_way <- mtcars %>% dplyr::filter(cyl == 6) 

# Compare the results 
head(base_r_way) # head() shows the first rows of a dataset
head(tidyverse_way)

```

### Data Summaries

Calculate the average mpg by cylinder:

```{r}
# Base R approach 
aggregate(mpg ~ cyl, data = mtcars, FUN = mean) 

```

```{r}
# Tidyverse approach 
mtcars %>% group_by(cyl) %>% summarise(avg_mpg = mean(mpg))

```

Reflection Questions (2 minutes)

What differences do you notice between tidyverse and base R syntax?
Which approach seems more readable to you and why? Can you identify
situations where one approach might be preferable over the other?

Key Takeaways The tidyverse is a collection of packages that work
together cohesively. Keep in mind possible function naming conflicts The
tidyverse approach often emphasizes readability

Additional Resources R for Data Science (free online):
<https://r4ds.had.co.nz/> Tidyverse website:
<https://www.tidyverse.org/> RStudio's cheatsheets:
<https://www.rstudio.com/resources/cheatsheets/>
