---
title: "Exercise 1.3: Finding Errors in Scripts and Functions"
author: "David Souto"
date: "`r Sys.Date()`"
output: html_document
---

## Importance of code readability

In this exercise, we will explore the importance of robust coding practices in R by identifying and fixing errors in a sample script. The provided code is poorly organized and contains an error that is not immediately obvious. Can you spot it?

Improve the organization of the code and naming so you can find the error upon running the chunk and be more explicit about what it does.

```{r sample-script, echo=TRUE}

# this is how functions work in R
# Define a function to calculate the square of a number
# Execute with Ctrl+Enter (running the first line will run the entire function); try it out with square(2)
square <- function(x) {
  result <- x^2  # Calculate the square
  return(result)  # Return the result
}

# Here is another function that generates an error 
fun <- function(numbers) { total <- sum(numbers); average <- total / length(numbers) return(average)}

```

# Evaluate style complinance 

We can use the `lintr` package to evaluate the style compliance of our R code. This package helps identify potential issues and improve code readability.

```{r}
# this package has a vignette, accessed with vignette("linter"), but no help
install.packages("lintr") 
library("lintr")

# run linter on this file (or any R file):
lint(filename="./ex1_3_finding_errors.Rmd")

# double click on the list of propositions to locate in the file, try to fix some of the errors, note that you will be on the "Markers" 

```

Key takeaway

Think about the readability of your code by using comments, code conventions and naming variables appropriately

# Level 2 (optional): Debugging tools

Imagine you want to calculate an average, 
Locating the unexpected source of an error can involve trial and error, locating the exact line that is problematic. 

Try to use browser() (and/or trace()) to locate errors within functions: https://rstudio-education.github.io/hopr/debug.html

To locate the source of the error, use the `browser()` function. Place `browser()` on the line just before what you suspect is the problematic code. When you run the function, execution will pause, allowing you to inspect the state of variables at that point.

Can you use browser to spot the error in the first chunk? Place a browser() on the line just before what you think is the culprit. 
Observe the extra ">" that appears on the console indicating you have paused at that location. Check the state of variables at that location
and run the erroneous line. Step out of the browser with "Esc".

Additional Resources

Narrated slides [here](https://utrechtuniversity.github.io/workshop-computational-reproducibility/chapters/readability.html) on code readability and lintr (R or python).

More exhaustive list of best [coding practice](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R.html) tips in R.
