---
title: "Exercise 4.3: Multiverse Analysis of Hurricane Names"
author: David Souto
output: html_document
---

In this exercise, we will explore the "female Hurricane" dataset to understand how different modeling choices can influence our results. This analysis will illustrate the concept of multiverse analysis, where we examine how varying our analytical choices can lead to different conclusions. Let's see how the femininity of hurricane names might affect the outcomes!

A complete analysis of this data has been done elsewhere; the analysis we are carrying out here is for illustration purposes. A more serious analysis would have needed to give more thought to the choice of model.

We are going to use the *specr* package to implement a basic multiverse analysis. Another option is *multiverse*.

Run the following chunk to make sure packages are installed and attach them

### Instructions
1. Set your working directory to the `day2-Challenges/01_session` folder where the data is located using with Session > Set Working Directory > Choose Directory.
2. Run the setup code to install and load the necessary packages.
3. Load the hurricane dataset and examine its structure.

```{r setup, include=FALSE}

# some options for chunks: we ask for it to produce outputs, no messages or warnings
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE) 

# You will bbe asksed whether you want to install from source, say yes
if (!require("specr")) 
  install.packages("specr") 

if (!require("tidyverse")) 
  install.packages("tidyverse") 

if (!require("ggplot2")) 
  install.packages("ggplot2") 

library(tidyverse)
library(specr)

```

### Load and Examine the Dataset

Load the hurricane dataset. Display the first few rows of the dataset and check its structure.

```{r , include=FALSE}

# we are loading the data-set, it takes the form of an .RData file, saved with save(), something
# that can be loaded straight away into R, after loading you will see data.frame object in the  Environment window.
load(file='./data/hurricane.RData') # this is where it helps to check the name of the data that was just loaded in the Environment window

# Display the first few rows of the dataset
head(hurricane_data)  # Check the structure of the dataset
str(hurricane_data)
```


How might the variables you've chosen influence the results of your analysis?

### Setting Up Specifications

```{r , include=FALSE}

# We are going to focus on the following variables in the Hurricane dataset
# death: number of deaths
# masfem: whether the name is seen masculine or feminine (rating scale)
# dam: damage
# zpressure: z-transformed atmospheric pressure

# This is more or less the original analysis in the Paper, showing a significant effect of name "feminity"
# We apply a general linear mode of a certain type (poisson)
# Poisson models are common for count data with a certain type of distribution (e.g., number of deaths, number of neurons action potentials)
fit <- glm(death ~ masfem * dam + masfem * zpressure, data = hurricane_data, family = "poisson")

# As odd as it sounds, we see a significant effect of the femininity of the hurricane (masfem)
summary(fit)

```
There are various effects, including an effect of femininity over mortality

### Execute Specification Curve Analysis

Now we are going to use functions within the specr package to specify different options for our analyses.

```{r}
specs <- setup(data = hurricane_data, # our dataset
   x = c("masfem", "masfem:dam", "masfem:zpressure"), # we are entering options for independent variables, those are our main variables of interest
   y = c("death","dam"), # options for dependent variables, we could predict damage amount instead of deaths
   control = c("masfem:dam", "masfem:zpressure"), # options for control variables, that is other predictors in our model
   model = c("lm","glm"), # options for models
   simplify = TRUE)
```

### Running the Analysis

We can how run all those analyses, amounting to 14 specifications.

```{r}
results <- specr(specs) # Summarize results summary(results)

```

### Explore the different specifications 

```{r}
summary(results)

```

### Visualization

After running the analysis, visualize the results to understand the impact of different specifications on predicting deaths.

```{r}
plot(results)

```

What patterns do you observe in the plot?

We coud sort the data by estimate with the dplyr function *arrange()* (as it is done in panel A of the figure above) so we can see which specifications yield the larger esetimates.

```{r}
results$data %>% 
  dplyr::arrange(estimate) %>% # sort by estimate
  dplyr::select(term,estimate,model,controls,formula) # select the columns you want to see     

```

Note there are many options we didn't explore yet. This particular example didn't consider the type of model that was used in the paper (poisson), for illustrative purposes.

Key Takeaways

The multiverse analysis allows researchers to explore multiple analytical paths and decision points, providing a more comprehensive view of how different choices can affect research outcomes.

The analysis of hurricane names illustrates how varying modeling choices—such as the inclusion of outliers, choice of dependent variables, and model types—can lead to different conclusions about the same dataset.

By making all analytical decisions explicit, multiverse analysis enhances transparency and helps identify potential biases or false positives in research findings.

Additional resources

For a more sophisticated analysis, check out the *multiverse* package and its application in the analysis of female hurricanes [here](https://rdrr.io/cran/multiverse/f/vignettes/example-hurricane.Rmd). This is fairly sophisticated as it uses custom code chunks.


