---
title: "Exercise 6.2:  Sequential testing with Bayes Factors"
author: David Souto
output: html_document
---

See the html version of this document to see the maths right.

# Introduction

Bayes factors are a powerful tool in Bayesian statistics that allow us to compare the evidence for two competing hypotheses or models. Unlike traditional null hypothesis significance testing (NHST), Bayes factors provide a continuous measure of evidence and can support both the null and alternative hypotheses.

The point of this worksheet is not to be a Bayesian statistics primer but to point to how we can use Bayes factors to plan our sampling. Because Bayes Factors can provide evidence in favour or against the null hypothesis we can use this a criterion to stop collecting data.

# What Are Bayes Factors?

The **Bayes factor (BF)** is the ratio of the marginal likelihoods of two competing models or hypotheses, \(M_1\) and \(M_2\). It quantifies how much more likely the observed data are under one model compared to the other. Mathematically, the Bayes factor is defined as:

$$
BF_{10} = \frac{P(D \mid M_1)}{P(D \mid M_2)}
$$

Where:
- \(P(D \mid M_1)\): The probability of the data \(D\) given model \(M_1\) (the marginal likelihood of \(M_1\)).
- \(P(D \mid M_2)\): The probability of the data \(D\) given model \(M_2\) (the marginal likelihood of \(M_2\)).

# Key Features of Bayes Factors

1. **Evidence for Both Hypotheses**: Unlike p-values, Bayes factors can provide evidence in favor of the null hypothesis.
2. **Continuous Evidence**: Bayes factors provide a continuous measure of evidence, rather than a binary reject/do-not-reject decision.
3. **Model Comparison**: Bayes factors inherently penalize more complex models, helping to avoid overfitting.

# Interpreting Bayes Factors

The strength of evidence provided by a Bayes factor can be interpreted using scales such as Jeffreys' scale or Kass and Raftery's scale. For example:

- \(BF_{10} > 10\): Strong evidence for \(M_1\).
- \(BF_{10} < 1/10\): Strong evidence for \(M_2\).
- \(BF_{10} \approx 1\): No evidence for either model.

# Sequential Testing with Bayes Factors

Sequential testing involves analyzing data as it is collected, rather than waiting for a fixed sample size. Bayes factors are particularly well-suited for sequential testing because they allow for optional stopping: You can stop data collection as soon as sufficient evidence is obtained.

This does not inflate Type I error rates, unlike NHST.

In sequential testing, Bayes factors are computed after each new data point (or batch of data) is collected. Data collection continues until the Bayes factor reaches a predefined threshold, indicating strong evidence for one hypothesis over the other.

# Simulating Bayes Factors for Sequential Testing

This chunk simulates sequential testing using Bayes factors for a comparison of two means.

```{r}
# Load required package
if (!requireNamespace("BayesFactor", quietly = TRUE)) {
  install.packages("BayesFactor")
}
library(BayesFactor)

# Simulate data for two groups
n_observations <- 100  # Sample size per group
group1 <- rnorm(n_observations, mean = 0, sd = 1)  # Group 1: Mean = 0
group2 <- rnorm(n_observations, mean = 0.5, sd = 1)  # Group 2: Mean = 0.5

# Sequential testing
threshold <- 10  # Bayes factor threshold for strong evidence
bf_values <- numeric(n_observations)  # Store Bayes factors

# we start at 2, we need at least 2 observations
for (i in 2:(n_observations)) {

  grp1 <- group1[1:i]
  grp2 <- group2[1:i]
  
  # Compute Bayes factor
  bf <- ttestBF(grp1,grp2, paired=FALSE)  # Compute Bayes factor
  
  # Check if bf is a valid object
  if (is.null(bf)) {
    stop("Bayes factor computation returned NULL.")
  }
  
  bf_values[i] <- 1/exp(bf@bayesFactor$bf)  # Extract BF10 (evidence in favour of hypothesis)
}

# Plot Bayes factors over time
plot(1:(n_observations), bf_values, type = "l", col = "blue", lwd = 2,
     xlab = "Sample Size per Group", ylab = "Bayes Factor (BF10)",
     main = "Sequential Testing with Bayes Factors")
abline(h = threshold, col = "red", lty = 2)  # Add threshold line

```

Reflection Questions:

How does the Bayes factor evolve as more data is collected?

Compare bayes factors with larger and smaller effects.

# Comparing Sequential Testing to NHST

This chunk compares the results of sequential testing with Bayes factors to a traditional t-test.

```{r}

# Perform a traditional t-test
t_test <- t.test(group1, group2)

# Final Bayes factor
final_bf <- ttestBF(group1, group2)

# Print results
cat("Traditional t-test p-value:", t_test$p.value, "\n")
cat("Final Bayes Factor (BF10):", exp(final_bf@bayesFactor$bf), "\n")

```
Reflection Questions:

How do the conclusions from the Bayes factor and the p-value compare?

What will happen when there is no effect?  

# Planning Sequential Testing with Bayes Factors

This chunk demonstrates how to plan a sequential testing study by simulating the expected sample size required to reach a decision depending on the effect size. We can see the benefit of this method, where we are are stopping when no more evidence is needed.

```{r}
set.seed(42)

# Function to simulate sequential testing
simulate_sequential_fun <- function(effect_size, n_max, threshold) {
  
  group1 <- rnorm(n_max, mean = 0, sd = 1)
  group2 <- rnorm(n_max, mean = effect_size, sd = 1)
  
  # need at least 2 obs
  for (i in 2:(n_max)) {
    grp1 = group1[1:i]
    grp2 = group2[1:i]
      
    bf <- ttestBF(grp1, grp2, paired=FALSE)
    
    if (exp(bf@bayesFactor$bf) > threshold) {
      return(i * 2)  # Return total sample size
    }
  }
  return(n_max)  # Return max sample size if threshold not reached
}

# Simulate for different effect sizes
effect_sizes <- seq(0.2, 0.8, by = 0.2)

# we use sapply to apply a function
sample_sizes <- sapply(effect_sizes, simulate_sequential_fun, n_max = 100, threshold = 10)

# Plot results
plot(effect_sizes, sample_sizes, type = "b", col = "blue", lwd = 2,
     xlab = "Effect Size", ylab = "Sample Size",
     main = "Sample Size vs. Effect Size in Sequential Testing")
```

Reflection Questions:

How does the required sample size change with the effect size?

Why is sequential testing more efficient for larger effect sizes?

Key Takeaways

Bayes Factors: Provide a continuous measure of evidence for or against hypotheses, unlike p-values.

Sequential Testing: Bayes factors allow for optional stopping without inflating Type I error rates, making them ideal for sequential testing. However, this stopping rule has to be specified in advance.

Efficiency: Sequential testing with Bayes factors often requires fewer participants than traditional NHST while maintaining robust conclusions.

Dienes Paper: https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2014.00781/full

"Thus, the following year another group of participants were run. One cannot simply top up participants with orthodox statistics, unless pre-specified as possible by one’s stopping rule (Armitage et al., 1969); by contrast, with Bayes, one can always collect more participants until the data are sensitive enough, that is, B < 1/3 or B > 3; see e.g., Berger and Wolpert (1988), Dienes (2008, 2011). Of course, B is susceptible to random fluctuations up and down; why cannot one capitalize on these and stop when the fluctuations favor a desired result? For example, Sanborn and Hills (2014) and Yu et al. (2014) show that if the null is true, stopping when B > 3 (if that ever occurs) increases the proportion of cases that B > 3 when the null is true. However, as Rouder (2014) shows, it also increases the proportion of cases that B > 3 when Ho is false, and to exactly the same extent: B retains its meaning as relative strength of evidence, regardless of stopping rule (for more discussion, see Dienes, forthcoming)."


Additional resources

Rouder https://pubmed.ncbi.nlm.nih.gov/24659049/

Some more technical discussion / debate: https://pure.rug.nl/ws/portalfiles/portal/210953159/Tendeiro2022_Article_Worked_outExamplesOfTheAdequac.pdf
