---
title: "Exercise 2.2: Using pipes to manipulate datasets"
output: html_document
---

We have seen pipes before, the `%>%` symbol that re-routes (think plumbing) the output of a function to another function. We are going to explore their use more systematically in the context of the tydiverse. Note there is another type of pipe syntax: |> but you might find this very awkward on your keyboard.

We are mainly going to use `dplyr` and `tidyr` functions and pipes to manipulate our data.

```{r}
# Run this chunk unless you have run Exercise 2.1
# The working directory needs setting to day1/02_session 

library(tidyverse)

tt_data <- read.csv("./data/tiktok_data.csv")

# make numbers variables numeric
tt_data$video_comment_count<-as.numeric(tt_data$video_comment_count)
tt_data$video_view_count<-as.numeric(tt_data$video_view_count)
tt_data$video_like_count<-as.numeric(tt_data$video_like_count)
tt_data$video_download_count<-as.numeric(tt_data$video_download_count)

# you will see a warning as indeed some of the data is NA and can't be coerced into a numeric value
```
# Data Cleaning

Data cleaning may involve removing NA values, stripping spaces, or converting data types.

# Remove rows with NA values

For some analysis, we could remove data for which we have incomplete information. With our data set it would 
correspond to removing rows whenever an NA value (not-a-number) is observed. 

```{r}
#  tt_data is routed to drop_na() which strips out rows containing NA occurrences, not-a-number
#  the result of that is assigned to tt_data
tt_data <- tt_data %>%
  drop_na()

# Now write and execute the equivalent in base R
# check the result (hint: head(), or as_tibble())
```

## Data manipulation functions in the tidyverse 

In the next chunk We are going to review `dplyr` functions `mutate`, `across`, and `where` to locate and manipulate data. The `dplyr` package has numerous options that we are not going to detail exhaustively.

# Use of mutate()

```{r}

# we are creating a new variable in the dataset, called long, which 
# takes a value of TRUE if video_duration is below 20 seconds, FALSE if above
tt_data <- tt_data %>%
  mutate(short = video_duration_sec>20) 

head(tt_data$short)

# Note the equivalence to: dplyr::mutate(tt_data, short=video_duration_sec>20)

# You can also specify the value for observations that meet the condition with
# if_else(), a dplyr function: if_else(video_duration_sec>20, "Short", "Long")

# create a new variable that takes the log of the video_comment_count, transform to integer if necessary

```

# Use of select() and filter() 

Our dataset is quite big, we may want to select specific variables to simplify things

```{r}

# Use of select()

# We could start by selecting only the numeric variables and the column verified_status with select() 
tt_data2<-tt_data %>% 
  # is.numeric() tells you whether a variable is numeric, whereas as.numeric is used to change the type
  select(is.numeric, verified_status)

# Try selecting by variable name

# Use of filter()

# We are interested in looking at the distribution of view counts, likes and shares

# let's see the frequency of videos downloads numbers 
# We can use base R to create a histogram of video download counts with specified x-axis limits
# We see most videos have z downloads
hist(tt_data2$video_download_count, 
     breaks = 5000,                # Number of bins
     xlim = c(0, 100),           # Set x-axis limits (adjust as needed)
     #ylim = c(0, 20),           # Set x-axis limits (adjust as needed)
     main = "Histogram of Video Download Counts",  # Title of the histogram
     xlab = "Video download Count",    # Label for the x-axis
     ylab = "Frequency",            # Label for the y-axis
     col = "lightblue",             # Color of the bars
     border = "black")              # Color of the bar borders

# We may be interested in filtering out videos with very few view counts
tt_data_f1 <- tt_data2 %>% 
  filter(video_view_count>1000)

# now filter out by video count AND filtering out accounts that were banned (author_ban_status) with '!=" (not equal to)
tt_data_f2 <- tt_data2 %>% 
  filter(_______, ______)

# run again the chunk but with the filtered data

```

# Produce summaries

We can use `group_by()` and `summarise()` to produce summaries of our data. Here we are going to count the number of posts by user status.

```{r}
counts_by_status <- tt_data %>%
  # we group by ban status
  group_by(author_ban_status) %>%
  # we summarize by status, and calculate the number of posts per group
  # note that n() does a count of rows within a group, it doesn't need to know the column
  summarise(post_count = n()) 

# use other type of summaries, using mean(), median(); and change the name of summary variables cacordingly. 
# Note that this time we need to know what (the column / variable) we are averaging

# View the results
print(counts_by_status)

# Check the average number of views per verified status 

```

# Combine functions

We can now use a combination of these functions to clean up, select and summarize data in a readable way.

```{r}

# Fill the blanks in the following chunk & execute

# Combine functions to summarize and filter data 
summary_combined <- tt_data2 %>% 
  # filter out videso with low view counts
  filter(video_view_count>20) %>% 
  # Select two variables 
  select(author_ban_status, video_view_count) %>% 
  # Group by author_ban_status
  group_by(author_ban_status) %>% 
  # provide summaries for view counts
  summarise( total_posts = ____, average_views = _____, max_views = ____ , min_views = ____)) 


```

Key Takeaways

Summarizing Data: The summarise() function allows you to compute various statistics like count, mean, and median for grouped data.

Combining Functions: You can chain multiple dplyr functions together to filter, select, and summarize data in a clear and concise manner.

Additional Resources

We only have touched upon most used functions of `dplyr`. More on [dplyr here](https://dplyr.tidyverse.org/).

The tiktok dataset can be access in [Kaggle.com](https://www.kaggle.com/datasets/yakhyojon/tiktok/data). It is a popular website among the 
data science / machine learning community. By selecting the "Code" tab you can see what multiple people have done with the dataset.


You might find the dplyr and tidyverse [cheatsheets useful](https://rstudio.com/resources/cheatsheets/).
