---
title: "Exercise 2.1: Loading datasets"
output: html_document
---

In this exercise, we will explore how to load and examine different types of datasets from various sources.

We are using a [Tiktok user engagement dataset](https://www.kaggle.com/datasets/yakhyojon/tiktok?resource=download) shows views and likes for a large number of videos.

R supports various data formats through specialized packages. With base R we can easily load .csv data or data that lives online.

```{r}
# Load necessary libraries and set the working directory to the correct session number
library(tidyverse)  
# set the working directory to day1/02_session 

tt_data <- read.csv("./data/tiktok_data.csv")

# We can also load a dataset from an URL: for instance, this letter frequency dataset in the english language
url <- "https://people.sc.fsu.edu/~jburkardt/data/csv/letter_frequency.csv"
freq_data <- read.csv(url)

```

# Display dataset information

Common `classes` for data are character,`chr`, numeric, `num`, which allows whole numbers and decimal numbers and integer, `int`. Note how numbers are at times stored as characters (between `""`) and therefore number operations cannot be applied to them.

```{r}
# Look at the tiktok data with print(), compare to the output in the Console
print(tt_data)

# Compare to just displaying the first 10 rows
head(tt_data)

# Check the properties of the variables
str(tt_data)

# you can also find the class

# Retrieve the names of the variables with names()

# Apply this to explore the letter frequency dataset 

# try using the console to see the output
```

# Factor type

In R, a factor is a data structure used to represent categorical data. Factors are particularly useful for statistical modeling and data analysis because they allow R to treat categorical variables appropriately. Here’s a detailed explanation of factors in R:

Levels: Each unique value in a factor is called a level. For instance, if you have a factor representing the colors of cars, the levels might be "red," "blue," and "green."

```{r}

# Change the letter variable of the word frequency dataset into a factor using as.factor()

# Print the result with print() or run in the Console

```

# Change data types and save data

You may want to save the dataset with the new datatypes.

```{r}

# Example: transform to integer the number of comments
tt_data$video_comment_count <- as.integer(tt_data$video_comment_count)

# check what happens
str(tt_data)

# do the same for likes and views

# change video_id to a character, using an analogue syntax to as.integer()

# you can save data pretty much in the same way you load data, using write.csv()
# use ?write.csv() to find out how to save the data in the same folder, appending v2 to the name.

# Load the version 2 of the dataset again and check whether the types have changed with str()

```

# Other useful dataset types: Tibbles

Tibbles are data frame types that are part of dplyr.

```{r}

# Observe the difference in displaying a dataset, when typing the following in the Console
head(as_tibble(tt_data)) # or more explicitly dplyr::as_tibble()

head(tt_data)

```

# Key Takeaways

Always check data types and structure after loading

Handle missing values appropriately

# Additional resources

Analyzing `big data`, data that would exceed memory capacity, is another kettle of fish. There are techniques and packages that enable handling of big data, as explained [here](https://hbs-rcs.github.io/large_data_in_R/).
