---
title: "Exercise 5.3: Power calculation with complex models through simulation"
author: David Souto
output: html_document
---

# Introduction

Power analysis for complex models, such as Generalized Linear Mixed Models (GLMMs), is more challenging than for simpler models like t-tests or ANOVAs. GLMMs account for both fixed and random effects, making them ideal for hierarchical or nested data, but this complexity also makes it difficult to compute power analytically.

In this worksheet, we’ll explore simulation-based power analysis for GLMMs using the simr package. By the end of this exercise, you’ll understand how to:
- Perform power analysis for GLMMs based on simulations.
- Explore the impact of sample size and effect size on power.
- Use simulation to plan studies with complex designs.

Run this chunk first to load the packages.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# we need simr and lme4 in this worksheet
if(!requireNamespace("simr", quietly = TRUE)){ 
    install.packages("simr")
}else{
    library("simr")
}
if(!requireNamespace("lme4", quietly = TRUE)){ 
    install.packages("lme4")
}else{
    library("lme4")
}
```


# Setting up a general linear mixed model (GLMM)

A General Linear Mixed Model is an extension of a General Linear Model incorporating fixed effects (e.g. treatment) and random effects (e.g. participants, schools, hospitals), and non-linear effects. It is useful for complex data structures, such as involving repeated measures, or hierarchical data, etc.  

We'll be using the example from the `simr` [vignette](https://cran.r-project.org/web/packages/simr/vignettes/examples.html)

Example Scenario

Imagine we are studying the effect of a treatment (in the broad sense) on a binary outcome (success or failure, hence we use a binomial model), modelling random effects on another factor. This could be the hospitals or patients, for instance. We hypothesize that the treatment increases the probability of success.

# Simulate a realistic dataset

Up to know we have simulated the simplest scenario. Check the following chunk line by line, where we simulate a realistic data in which there is an effect of treatment and an effect of hospital on a health outcome. Note that we have a single value per patient, so we don't need to model patient as a random effect.


```{r}

# Set seed for reproducibility
set.seed(42)

# Define parameters
n_patients <- 20          # Number of patients
obs_per_patient <- 5      # Number of observations per patient (can vary)
total_obs <- n_patients * obs_per_patient  # Total number of observations

# Generate patient IDs
patient_id <- rep(1:n_patients, each = obs_per_patient)

# Generate random effects for patients
patient_effect <- rnorm(n_patients, mean = 0, sd = 2)

# Assign treatments (randomly)
treatment <- rep(c(0, 1), length.out = total_obs)
treatment <- sample(treatment)  # Randomize treatment assignment

treatment_effect = 1 # by how much the score differs when we apply the treatment

# Generate repeated measures (observations per patient)
health_score <- 50 + 
                treatment_effect * treatment +      # Treatment effect
                patient_effect[patient_id] +       # Random patient effects
                rnorm(total_obs, mean = 0, sd = 2) # Residual error

# Create dataset
data <- data.frame(
    patient = factor(patient_id),
    treatment = factor(treatment, labels = c("Control", "Treatment")),
    health_score = health_score
)

```

Now we can model the effect of treatment and hospital. 

```{r}
model <- lmer(health_score ~ treatment + (1 | patient), data = data) 

# Summary of the model 
summary(model)
```
Questions:

What is the estimated effect of the treatment on success?

Why do we include (1 | patient) in the model?

# Power Analysis for the GLMM

We’ll use the `simr` package to estimate power by simulating data and testing for significance. Fill the blanks and run the following code to calculate power for the treatment effect:

```{r}
# We can now perform a power analysis with powerSim (part of simr)
# deciding on running 100 simulations (we'll run more if we wanted a more accurate estimate)
power <- powerSim(model, nsim = 100) # Run 100 simulations 

print(power)

# Fill the blank, increase the number of patients and see what the power is
model_with_more_participants <- extend(model, along = "patient", n = __) 
power <- powerSim(model_with_more_participants, nsim = 100) # Run 100 simulations 

print(power)

```
What is the estimated power for detecting the treatment effect?

How does increasing the number of participants affect power?

# Power Curve for Sample Size

Let’s explore how power changes with the number of patients.

```{r}
# Fill the blank Define a range of group sizes
# the maximum has to be smaller or equal to the number of patients you picked above
group_sizes <- seq(__,__, by = 5) 

# Calculate power for each group size (this takes a bit longer, we'll reduce the number of simulations)
power_curve <- powerCurve(model_with_more_participants, along = "patient", breaks = group_sizes, nsim = 20) 

# Plot the power curve 
plot(power_curve)
```

Questions

At what group size does power reach 80%?

# Exploring the Impact of Effect Size

Modify the code below to calculate power for different treatment effect sizes (( \beta = 0.2, 0.4, 0.6 )):

```{r}

# We can also adjust the effect size (non standardized) with fixef()
fixef(model)["treatmentTreatment"]<-0.4

# Change the treatment effect size 
# Perform power analysis 
power_effect <- powerSim(model, nsim = 100) 

print(power_effect)

```

# Key Takeaways

You can now use simulation to generate power calculations that incorporate your knowledge of the data in complex models, as allowed by GLMM. This can include knowledge regarding the reliability of the measures (which we didn't address).

This can be used in addition to analytically solutions with tools such as G*Power.

# Additional Resources

Very detailed tutorial expanding on different situations, such as how we can improve power by changing the design and how to deal with interactions: 

Brysbaert, M. (2019). How many participants do we have to include in properly powered experiments? A tutorial of power analysis with reference tables. Journal of cognition, 2(1), 16. https://pmc.ncbi.nlm.nih.gov/articles/PMC6640316/
